NumPy---Array Computation for Python

### 2019-12-17 paper.tex
NEW: Start on contributions section and create plot/script for that
Note that the pickle data is not included here, because it contains
semi-private data (although it probably does not matter), plus it
is a bit large at 1.5MB

% New Commands
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\fixme}[1]{\textcolor{red}{{#1}}}
\newcommand{\inlinecite}[1]{\footnotesize\citen{#1}}

@@

\subsection*{\_\_array\_func\_\_}

@@

>>>>

% New Commands
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\fixme}[1]{\textcolor{red}{{#1}}}
\newcommand{\inlinecite}[1]{\footnotesize\citen{#1}}
\newcommand{\tightlist}{\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

\subsection*{The \_\_array\_function\_\_ Protocol}

The central position which NumPy takes within the python ecosystem with
respect to scientific computing and numerical works,
means that a vast number of projects are build on it.
These projects are consumers of the NumPy API.

On the other side, projects are providers of a
\emph{NumPy-like API} and array objects targeting audiences with specialized needs beyond
NumPy's capabilities. Some of these projects provide:

\begin{description}
    \tightlist
    \item[SciPy's \code{scipy.sparse} and \code{pydata.sparse}:] Sparse arrays
    \item[CuPy:] Computation on the GPU
    \item[Dask:] Parallel, distributed computing
    \item[PyTorch and TensorFlow:] Deep learning frameworks.
\end{description}

The multitude of specialized projects creates the difficulty that consumers
of these NumPy-like APIs write code specific to a single project and do not support
all of the above array providers.
This is a burden for users relying on the specialized Array-like, since
a tool they need may not work for them.
Additionally, a user who has the need of transitioning from NumPy to a more
specialized array, is faced with a challenge.
These issue can lead into a fragmentation of the user bases into consumers
of different NumPy-like APIs.

To address these issues NumPy has the goal of providing the fundamental
API for \emph{interoperability} between the various NumPy-like APIs.
An earlier step in this direction was the implementation of the
\code{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
for most mathematical functions.\cite{NEP13}
In 2019 this was expanded more generally with the inclusion of the
\code{\_\_array\_function\_\_} protocol into NumPy~1.17.
These two protocols allows providers of Array-like APIs to be interoperable
with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
For the users relying on specialized array projects it means that even though
much code is written specifically for NumPy arrays and uses the NumPy API as
\code{import numpy as np}, it can nevertheless work for them.
A brief example of such a function which is written using NumPy
but can now be used more generally is:
\begin{lstlisting}
def f(x):
    y = np.tensordot(x, x.T)
    return np.mean(np.exp(y))
\end{lstlisting}
For example, a user relying on this function and normally using it with a
NumPy array, may now use a distributed Dask array when challenged
by a very large dataset, without any modification to \code{f(x)}.

While, as of writing this, the interface is still considered experimental
and may be modified, it has been implemented by multiple projects already
and proven useful.
% TODO: Should likely have citation/example if I claim it was useful (based
%       mainly on Matt's statement).

@@

NumPy is currently maintained a group of 23 contributors with commit rights
to the NumPy code base. Out of these 17 maintainers were active in 2019.
Additionally, there are a few long term developers who contributed and maintain
specific parts of NumPy, but are not officially maintainers.
At a release cycle of about every half year, the five recent releases in the years
2018 and 2019 have averaged about 450~pull requests~(PR) each.\footnote{
    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
    All data provided here is based on the development which happened through github
    pull requests. In some cases contributions by maintainers may not be categorized as such.}
% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
With each release attracting more than a hundred new contributors.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
    \caption{Number of pull requests merged into the NumPy master branch for each
        quarter since 2012. The total number of PR is indicated with the
        lower blue area showing the portion contributed by current or previous
        maintainers.}\label{fig:prs-over-time}
\end{figure}

The stacked plot in Figure~\ref{fig:prs-over-time} shows the number of
PR merged into the NumPy master branch.
In this plot PR opened by current and old maintainers are indicated.
Although the number of PRs being merged fluctuates,
the plot indicates an increased number of contributions over the past
years.
Over the course of its history, NumPy has attracted PR by 823 contributors.
However, its development still relies heavily on the relative small number
of stable maintainers, who share more than half of the contributions among
themselves.
